Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-10 analysis/paper/paper.Rmd
EDITS PER SKH COMMENTS

The global climate crisis is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl; @meier2014rg; @wang2017dsrpitsio]. These
changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed, bearded, spotted, and ribbon seals) which depend on spring
and early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
important functions such as pupping, nursing, breeding, and molting [@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
ability to make definitive predictions about the ultimate effects of changes in sea ice on
the behavior, abundance, and distribution of these seals. Knowledge about biological
constraints on the phenology of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal species will be
able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). Additionally, trends in abundance of these species
are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals do
rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The very remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
bearded seals spent hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
primary focus was to develop a correction factor for estimating seal abundance, and like a
subsequent analysis by Conn et al. [-@conn2014mee], they did not attempt to separately
analyze haul-out schedules for different age and sex classes (although initial models with
sex and age as fixed effects were insignificant at p = 0.05; [@verhoef2014sm]).

@@

In this study, we use data from fifteen years of bio-logger records in a comprehensive
investigation of haul-out behavior by three ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
spring months and they warrant a separate study that includes additional environmental
covariates such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
were representative of the population as a whole, even though the age and sex distribution
of telemetered individuals might differ from the population. Third, we explore the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability
and possible ramifications of non-representative age-sex samples, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt and limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young-of-the-year. subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on March 1 of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

deploy_tbl %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue("{n} ({seal_hours} seal hours)")
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young of Year` = `YOUNG OF YEAR`,
                `subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young of Year` = "Age Class",
   `subadult` = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

@@

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites and we did not want to include data from these tags in our analyses. As
such, start and end times of each deployment were identified by experts who examined several
simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
ceased providing data consistent with seal behavior. Sensor readings outside of these
start and end times were discarded prior to analysis. After approximately 9 months, a few
(7) devices deployed on the rear flipper of bearded seals reported implausible sensor
data. All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the
study season for each species. Observations are concentrated in the months of May and June
due to the timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the water and
more data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
record. Tags were deployed and programmed in a manner to maximize data transmission during
the spring pupping and molting period. However, not all days during the deployment were
successfully transmitted. This is due to a variety of factors including satellite
coverage, tag availability (i.e. tags mounted to the rear flipper are not available to the
surface while the seal is at sea), tag performance, duty cycling, and extra-terrestrial
interference.

@@

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a movement
model to predict locations at specific times. However, the sparse nature of data from some
tags, especially those mounted to the rear flipper, resulted in poor modeling performance
or convergence issues. For this study, we calculated a weighted average daily location
where the inverse of the estimated Argos location error was used for the weight. Any days
where haul-out observations were present without any location data were filled from the
last calculated daily location and any days where the location intersected with land were
removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
available haul-out behavior data for each species across the study area.

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, and wind chill (represented
by a wind $\times$ temperature interaction; [@udevitz2009cjz]), together with day-of-year
and time of day interactions to permit diurnal patterns to change throughout the year).
Spotted and ribbon seal models included age-sex class and interactions between day-of-year
and age-sex class, but we omitted these from bearded seal models due to poor
representation of age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models did
include a latitudinal effect (and an interaction with day-of-year) since bearded seals
occupy a substantial range and we were interested in possible differences in the timing of
haul-out behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models since, during the spring, these species are most prevalent
near the southern ice edge in the Bering Sea [@conn2014mee].

```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 2012 May 05. The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice. Seals haul out in the presence of sea ice and we could assess the local concentration
of sea ice during these events. However, seals also spend time in the water away from sea
ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
ice concentration. Would it be the concentration at the last known haul-out event? The
next known haul-out event? Or, the closest sea ice concentration available? Without a
meaningful means to associate a sea ice covariate with the seal throughout the duration of
the observation record, such parameters are not viable for inclusion in this model.

We did explore whether the annual variation in maximum spring sea ice extent in the Bering
Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get the total sea ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily value for each year.

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment . Owing to the large number of records,
we fit separate models to bearded, ribbon, and spotted seal data sets. We fit two
different models to our datasets. First, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted seals (which
had considerably more data than bearded seals), we fitted models that included all the
effects from the first model, but also permitted annual variation in haul-out timing. This
second class of models was used to examine whether haul-out patterns varied by year, and
to determine the annual timing of apparent peaks in haul-out behavior.

@@

To determine how much of a potential problem this is for our data, we compared estimates
of haul-out probabilities that (1) ignored sex- and age-structure, and (2) accounted for
sex- and age-structure, adjusting for the relative proportion of seals in each sex- and
age-class. To simplify comparisons, we fitted two additional GLMPMs to ribbon and spotted
seal haul-out datasets. In the first model, we expressed haul-out probabilities as a
function of solar hour and day-of-year. In the second, we also included effects of sex-age
class and sex-age interactions with solar hour and day-of-year. To adjust for the
proportion of seals in each age class, we calculated a predicted population-averaged
haul-out probability at solar noon as a function of day-of-year ($t$) as

@@

To calculate plausible values for $\omega_{s,a}$ we obtained natural mortality predictions
from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young-of-the-year: 0-1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

@@

The haul-out probability for ribbon seals was strongly influenced by each of the weather
covariates included within the model except precipitation. The impact of temperature
($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$ = 
`r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`), wind
($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$ = 
`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric pressure
($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ = 
`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) was especially
noticeable as ribbon seals were less likely to haul out at higher winds and lower pressure
values and more likely to be on the ice when temperatures were relatively warm. Wind chill
($F_{`r paste(ribbon_type3$temp2_wind$Num.df, ribbon_type3$temp2_wind$Den.df, sep=",")`}$ 
= `r ribbon_type3$temp2_wind$F.value`; $p$ = `r ribbon_type3$temp2_wind$Prob.F`) showed a
moderately negative influence on haul-out probability.

@@

The second set of models that included annual variation in haul-out patterns uncovered
significant contributions for linear and quadratic interactions between day and year in
both ribbon seals (day:year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$ = 
`r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$ = 
`r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`) and
spotted seals (day:year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$ = 
`r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$ = 
`r spotted_year_type3$day2_year$F.value`; $p$ = `r spotted_year_type3$day2_year$Prob.F`).
Predicted distributions of haul-out activity were largely unimodal, but varied some among
and within years with respect to the timing and magnitude of haul-out peaks (Figure
\@ref(fig:annualHO)). It is important to note that predicted variation in annual haul-out
patterns likely reflects both process error and sampling variability. While we did remove
any years where only one deployment in a species + age-sex group was present, there are
still some years where the pattern shown is informed by a small number of tags that may
not represent population level processes.

```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 01-June or ended before 01-May are not included."}
select_groups <- fit_ribbon_year$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

>>>>

The global climate crisis is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl; @meier2014rg; @wang2017dsrpitsio]. These
changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed, bearded, spotted, and ribbon seals) that depend on spring
and early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
ability to make definitive predictions about the ultimate effects of changes in sea ice on
the behavior, abundance, and distribution of these seals. Knowledge about biological
constraints on the phenology of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal species will be
able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). Additionally, trends in abundance of these species
are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals
rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
bearded seals spent hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
primary focus was to develop a correction factor for estimating seal abundance, and like a
subsequent analysis by Conn et al. [-@conn2014mee], they did not attempt to separately
analyze haul-out schedules for different age and sex classes (although initial models with
sex and age as fixed effects were insignificant at p = 0.05; [@verhoef2014sm]).

@@

In this study, we use data from 15 years of bio-logger records in a comprehensive
investigation of haul-out behavior by 3 ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
spring months, and they warrant a separate study that includes additional environmental
covariates, such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
were representative of the population as a whole, even though the age and sex distribution
of telemetered individuals might differ from the population. Third, we explore the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability
and possible ramifications of non-representative age-sex samples, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young-of-the-year. Subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

deploy_tbl %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue("{n} ({seal_hours} seal hours)")
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young of Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young of Year` = "Age Class",
   `subadult` = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

@@

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites, and we did not want to include data from these tags in our analyses. As
such, start and end times of each deployment were identified by experts who examined
several simultaneous data streams (e.g., tag locations and dive behavior) to determine
when tags ceased providing data consistent with seal behavior. Sensor readings outside of
these start and end times were discarded prior to analysis. After approximately 9 months,
7 devices deployed on the rear flipper of bearded seals reported implausible sensor data.
All data after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the study
season for each species. Observations are concentrated in the months of May and June due
to the timing of deployment (April and May) and the timing of molt (May and June). During
molt, seals (and their attached bio-loggers) spend more time out of the water and more
data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
record. Tags were deployed and programmed in a manner to maximize data transmission during
the spring pupping and molting period, however, not all days during the deployment were
successfully transmitted. This is due to a variety of factors including satellite
coverage, tag availability (i.e. tags mounted to the rear flipper are not available to the
surface while the seal is at sea), tag performance, duty cycling, and extra-terrestrial
interference.

@@

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a movement
model to predict locations at specific times. However, the sparse nature of data from some
tags, especially those mounted to the rear flipper, resulted in poor modeling performance
or convergence issues. For this study, we calculated a weighted average daily location
where the inverse of the estimated Argos location error was used for the weight. Any days
where haul-out observations were present without any location data were filled from the
last calculated daily location, and any days where the location intersected with land were
removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
available haul-out behavior data for each species across the study area.

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill (represented
by a wind $\times$ temperature interaction [@udevitz2009cjz]), and day-of-year
and time of day interactions to permit diurnal patterns to change throughout the year.
Spotted and ribbon seal models included age-sex class and interactions between day-of-year
and age-sex class, but we omitted these from bearded seal models due to poor
representation of age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models
included a latitudinal effect (and an interaction with day-of-year), since bearded seals
occupy a substantial range and we were interested in possible differences in the timing of
haul-out behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models since, during the spring, these species are most prevalent
near the southern ice edge in the Bering Sea [@conn2014mee].

```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012. The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice; seals haul out in the presence of sea ice and we could assess the local concentration
of sea ice during these events. However, seals also spend time in the water away from sea
ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
ice concentration. Would it be the concentration at the last known haul-out event? The
next known haul-out event? Or, the closest sea ice concentration available? Without a
meaningful means to associate a sea ice covariate with the seal throughout the duration of
the observation record, such parameters are not viable for inclusion in this model.

We explored whether the annual variation in maximum spring sea ice extent in the Bering
Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get the total sea ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily value for each year.

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment . Owing to the large number of records,
we fit separate models to bearded, ribbon, and spotted seal data sets. We fit two
different models to our datasets. First, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted seals (which
had considerably more data than bearded seals), we fitted models that included all the
effects from the first model, but also permitted annual variation in haul-out timing. This
second class of models was used to examine whether haul-out patterns varied by year and
to determine the annual timing of apparent peaks in haul-out behavior.

@@

To determine how much of a potential problem this is within our data, we compared
estimates of haul-out probabilities that (1) ignored sex- and age-structure and (2)
accounted for sex- and age-structure, adjusting for the relative proportion of seals in
each sex- and age-class. To simplify comparisons, we fitted two additional GLMPMs to
ribbon and spotted seal haul-out datasets. In the first model, we expressed haul-out
probabilities as a function of solar hour and day-of-year. In the second model, we also
included effects of sex-age class and sex-age interactions with solar hour and
day-of-year. To adjust for the proportion of seals in each age class, we calculated a
predicted population-averaged haul-out probability at solar noon as a function of
day-of-year ($t$) as

@@

To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality predictions
from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young-of-the-year: <1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

@@

The haul-out probability for ribbon seals was strongly influenced by each of the weather
covariates included within the model except precipitation. The impacts of temperature
($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$ = 
`r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`), wind
($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$ = 
`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric pressure
($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ = 
`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) were especially
noticeable as ribbon seals were less likely to haul out at higher winds and lower pressure
values and more likely to be on the ice when temperatures were relatively warm. Wind chill
($F_{`r paste(ribbon_type3$temp2_wind$Num.df, ribbon_type3$temp2_wind$Den.df, sep=",")`}$ 
= `r ribbon_type3$temp2_wind$F.value`; $p$ = `r ribbon_type3$temp2_wind$Prob.F`) showed a
moderately negative influence on haul-out probability.

@@

The second set of models that included annual variation in haul-out patterns uncovered
significant contributions for linear and quadratic interactions between day and year in
both ribbon seals (day $\times$ year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$ = 
`r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$ = 
`r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`) and
spotted seals (day $\times$ year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$ = 
`r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$ = 
`r spotted_year_type3$day2_year$F.value`; $p$ = `r spotted_year_type3$day2_year$Prob.F`).
Predicted distributions of haul-out activity were largely unimodal, but varied some among
and within years with respect to the timing and magnitude of haul-out peaks (Figure
\@ref(fig:annualHO)). It is important to note that predicted variation in annual haul-out
patterns likely reflects both process error and sampling variability. While we did remove
any years where only one deployment in a species + age-sex group was present, there are
still some years where the pattern shown is informed by a small number of tags that may
not represent population level processes.

```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
select_groups <- fit_ribbon_year$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)
